When did a variant start spreading?

### 2022-07-23 notes/variant_origin.tex
Idea for Q computation in O(N^3k)

Define the $\textbf{m}_t = (m_0, ..., m_k)_t$ vector as the sampled number of genotypes with 0,1,..., k mutations at time $t$, then  $\textbf{n}_t = (n_0, ..., n_k)_t$ is the hidden true population size at each of these time points. Now the probability of the ${\textbf{m}_t}_t$ vector with seeding time of the first genotype at time $\tau$ (here the hidden vector is $\textbf{n}_\tau = (1, 0, ..., 0)$) can be expressed as

@@

This is due to the fact that populations of genotypes with 1 or more mutations are not only dependent on the size of that population in the previous time step but also on the size of the population with 1 less mutation (we assume that the likelihood of more than one mutation in a generation step is negligibly small).

\begin{equation}
g'_{n_{i,t},n_{i,t-1}, n_{i-1,t-1}} = g_{n_{i,t},n_{i,t-1}} + \mu L n_{i-1, t-1}
\end{equation}

@@

However, this also means that the probability is not comprised of linearly independent sub-probabilities for each genotype's observed samples. Thus, in order to calculate the maximum likelihood seeding time $R$ and $Q$ must be calculated by iterating over a k dimensional subspace.

@@

\end{document}

>>>>

Define the $\textbf{m}_t = (m_0, ..., m_k)_t$ vector as the sampled number of genotypes with 0,1,..., k mutations at time $t$, then  $\textbf{n}_t = (n_0, ..., n_k)_t$ is the hidden true population size at each of these time points. Now the probability of the ${\textbf{m}}_t$ vector with seeding time of the first genotype at time $\tau$ (here the hidden vector is $\textbf{n}_\tau = (1, 0, ..., 0)$) can be expressed as

@@

This is due to the fact that population sizes of genotypes with 1 or more mutations are not only dependent on the size of that population in the previous time step but also on the size of the population with 1 less mutation, i.e the likelihood a mutation happened in the last time step (we assume that the likelihood of more than one mutation in a generation step is negligibly small). Thus, we can describe each transition probability by the previously defined transition matrix and the mutation probability.

\begin{equation}
g'_{n_{i,t},n_{i,t-1}, n_{i-1,t-1}} = \sum_{\epsilon=0}^{n_{i,t}} g_{n_{i,t} - \epsilon,n_{i,t-1}} \cdot  \chi_{\epsilon, n_{i-1, t-1}}
\end{equation}

@@

However, this also means that the probability is not comprised of linearly independent sub-probabilities for each genotype's observed samples and the maximum likelihood seeding time $R$ and $Q$ are expressed by iterating over a k dimensional subspace.

@@

For many dimensions this becomes too complex to easily compute, therefore we utilize the fact that the transition matrix g is only maximally dependent on 2 sub-spaces. Note that $Q$ is equivalent to 
\begin{equation}
\begin{split}
    Q(t,\tau)_{\textbf{n}} &= P(\{m_i\}_{1 \leq i<t} | \textbf{n}_t = \textbf{n}, \tau) \\
    & = P(\{m_{0,i}\}_{1 \leq i<t} | n_{0,t} = n^0, \tau) P(\{m_{1,i}\}_{1 \leq i<t} | \{m_{0,i}\}_{1 \leq i<t}, n_{0,t} = n^0, n_{1,t} = n^1, \tau) \cdot ...\\ \cdot & P(\{m_{k,i}\}_{1 \leq i<t} | \{m_{k-1,i}\}_{1 \leq i<t}, n_{k-1, t} = n^{k-1}, n_{k,t} = n^k, \tau)
    \end{split}
\end{equation}

Now each of these terms can be calculated in $\mathcal{O}(N^3)$, starting with genotype 0 and moving up to genotype $k$.
\begin{equation}
\begin{split}
    P(\{m_{0,i}\}_{1 \leq i<t} | n_{0,t} = n^0) 
    & = \sum_{n'} g_{n^0, n'} P(m_{0, t-1} | n_{0, t-1} = n', \epsilon) P(\{m_{0,i}\}_{1 \leq i<(t-1)} | n_{0,t-1} = n') \\
    &= \sum_{n'} x_{0, t-1, n^0}(n') \\
    \\
    P(\{m_{1,i}\}_{1 \leq i<t} | \{m_{0,i}\}_{1 \leq i<t}, n_{1,t} = n^1, n_{0,t} = n^0) 
    & = \sum_{n_1'} P(m_{1, t-1} | n_{1, t-1} = n_1', \epsilon) \cdot \\
        & \sum_{n_0'}  g_{n^0, n_0'} P(m_{0, t-1} | n_{0, t-1} = n_0', \epsilon) P(\{m_{0,i}\}_{1 \leq i<(t-1)} | n_{0,t-1} = n_0') \cdot \\
        & g_{n^1, n_0', n_1'} P(\{m_{1,i}\}_{1 \leq i<(t-1)} | \{m_{0,i}\}_{1 \leq i<(t-1)}, n_{1,t-1} = n_1', n_{0,t-1} = n_0')  \\
    &= \sum_{n_1'} P(m_{1, t-1} | n_{1, t-1} = n_1', \epsilon) \sum_{n_0'} x_{0, t-1, n^0}(n_0') y_{0, 1, t-1, n^1}(n_0', n_1') \\
    &= \sum_{n_1'} P(m_{1, t-1} | n_{1, t-1} = n_1', \epsilon) x_{0, 1, t-1, n^0, n^1}(n_1') \\
    \\
    &... \\
    \\
    P(\{m_{k,i}\}_{1 \leq i<t} | \{m_{k-1,i}\}_{1 \leq i<t} n_{k,t} = n^k, n_{k-1,t} = n^{k-1}) 
    &= \sum_{n_k'} P(m_{k, t-1} | n_{k, t-1} = n_k', \epsilon) \cdot \\
    & \sum_{n_{k-1}'} y_{k-1, k, t-1, n^k}(n_{k-1}', n_k') \sum_{n^{k-2}} x_{k-2, k-1, t-1, n^{k-2}, n^{k-1}}(n_{k-1}') \\
    &= \sum_{n_k'} P(m_{k, t-1} | n_{k, t-1} = n_k', \epsilon)   x_{k-1, k, t-1, n^{k-1}, n^k}(n_k') \\
    \\
\end{split}
\end{equation}

We can perform the same operations to obtain $R$.
\begin{equation}
\begin{split}
    R(t)_{\textbf{n}} &= P(\{m_i\}_{t \leq i \leq T} | \textbf{n}_t = \textbf{n}) \\
    & = P(\{m_{0,i}\}_{t \leq i \leq T} | n_{0,t} = n^0) P(\{m_{1,i}\}_{t \leq i \leq T} | \{m_{0,i}\}_{t \leq i \leq T}, n_{0,t} = n^0, n_{1,t} = n^1) \cdot ...\\ \cdot & P(\{m_{k,i}\}_{t \leq i \leq T} | \{m_{k-1,i}\}_{t \leq i \leq T}, n_{k-1, t} = n^{k-1}, n_{k,t} = n^k)
    \end{split}
\end{equation}
\begin{equation}
    \begin{split}
P(\{m_{0,i}\}_{t \leq i \leq T} &| n_{0,t} = n^0) = P(m_{0,t} | n_{0,t} = n^0, \epsilon) \sum_{n'} g_{n', n^0}  P(\{m_{0,i}\}_{t+1 \leq i \leq T} | n_{0,t+1} = n')\\
&= P(m_{0,t} | n_{0,t} = n^0, \epsilon) \sum_{n'} x_{0, n^0}(n')
\\
P(\{m_{1,i}\}_{t \leq i \leq T} &| \{m_{0,i}\}_{t \leq i \leq T}, n_{0:1,t} = (n^0, n^1)) = P(m_{1,t} | n_{1,t} = n^1, \epsilon) P(\{m_{1,i}\}_{t+1 \leq i \leq T} | \{m_{0,i}\}_{t \leq i \leq T}, n_{0,t} = n^0, n_{1,t} = n^1) \\
&= P(m_{1,t} | n_{1,t} = n^1, \epsilon) \cdot\\
& \sum_{n_1'} \sum_{n_0'} P(\{m_{1,i}\}_{t+1 \leq i \leq T} | \{m_{0,i}\}_{t \leq i \leq T}, n_{0,t} = n^0, n_{1,t} = n^1, n_{1, t+1} = n_1', n_{0, t+1} = n_0')\\
&= P(m_{1,t} | n_{1,t} = n^1, \epsilon) \sum_{n_1'} \sum_{n_0'} g_{n_1', n^0, n^1} x_{0, n^0}(n_0') \cdot \\
&P(\{m_{1,i}\}_{t+1 \leq i \leq T} | \{m_{0,i}\}_{t+1 \leq i \leq T}, n_{1, t+1} = n_1', n_{0, t+1} = n_0')
\\
&= P(m_{1,t} | n_{1,t} = n^1, \epsilon) \sum_{n_1'} x_{0, 1, n^0, n^1}(n_1')
    \\
    &... \\
    \\
P(\{m_{k,i}\}_{t \leq i \leq T} &| \{m_{k-1,i}\}_{t \leq i \leq T}, n_{k-1:k,t}= (n^{k-1}, n^k)) 
= P(m_{k,t} | n_{k,t} = n^k, \epsilon) \sum_{n_k'} \sum_{n_{k-1}'} g_{n_{k-1}', n^{k-1}, n^k} \cdot \\
& \sum_{n_{k-2}'} x_{k-2, k-1, n_{k-2}', n^{k-1}}(n_{k-1}')
P(\{m_{k,i}\}_{t+1 \leq i \leq T} | \{m_{k-1,i}\}_{t+1 \leq i \leq T}, n_{k, t+1} = n_k', n_{k-1, t+1} = n_{k-1}')
\\
&= P(m_{k,t} | n_{k,t} = n^k, \epsilon) \sum_{n_k'} x_{k-1, k, n^{k-1}, n^k}(n_k')
    \end{split}
\end{equation}
\end{document}
